---
title: "Stat331 final project"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Setup

```{r}
library(glmnet)
library(corrplot)
pollutants_original <- read.csv("pollutants.csv", header = TRUE)
pollutants <- pollutants_original[-1]
```

```{r}
gender_list <- c('female', 'male')
edu_list <- c('before_high_school', 'high_school', 'college', 'college_grad')
eth_list <- c('other', 'mexi_us', 'nonhisp_black', 'nonhisp_white')
smoke_list <- c('no', 'yes')

gender <- gender_list[pollutants$male + 1]
education <- edu_list[pollutants$edu_cat]
race <- eth_list[pollutants$race_cat]
smoke_now <- smoke_list[pollutants$smokenow + 1]
```

```{r}
pollutants$male <- factor(gender)
pollutants$edu_cat <- factor(education)
pollutants$race_cat <- factor(race)
pollutants$smokenow <- factor(smoke_now)

pollutants$male <- factor(gender, levels=gender_list)
pollutants$edu_cat <- factor(education, levels=edu_list)
pollutants$race_cat <- factor(race, levels=eth_list)
pollutants$smokenow <- factor(smoke_now, levels=smoke_list)
```

```{r}
## Factors removed 
pollutants_factorsRemoved <- pollutants[-c(27,28,29,32)]

pollutants_factorsRemovedCor <- cor(pollutants_factorsRemoved)
```

Train and test data

```{r}
set.seed(57)

N <- nrow(pollutants)
sampleTrain <- sample(1:N, round(N*0.8,0), replace = FALSE)

dataTrain <- pollutants[sampleTrain,]
dataTest <- pollutants[-sampleTrain,]
```

Bounds for model selection

```{r}
M0 <- lm(length ~ 1, data = dataTrain)
Mfull <- lm(length ~ ., data = dataTrain)
```

```{r}
## Forward selection using AIC
MfwdAIC <- step(object = M0, scope = list(lower = M0, upper = Mfull), 
             trace = FALSE, direction = "forward", k = 2)
summary(MfwdAIC)

## Prediction accuracy
MfwdAIC.res <- pollutants$length[-sampleTrain] - predict(MfwdAIC, newdata = dataTest)
mspeMfwdAIC <- mean(MfwdAIC.res^2)

print(paste("MSPE of foward selection model based on AIC:",mspeMfwdAIC))
```

```{r}
## Forward selection using BIC
MfwdBIC <- step(object = M0, scope = list(lower = M0, upper = Mfull), 
             trace = FALSE, direction = "forward", k = log(nrow(dataTrain)))
summary(MfwdBIC)

## Prediction accuracy
MfwdBIC.res <- pollutants$length[-sampleTrain] - predict(MfwdBIC, newdata = dataTest)
mspeMfwdBIC <- mean(MfwdBIC.res^2)

print(paste("MSPE of foward selection model based on AIC:",mspeMfwdBIC))
```

```{r}
## Backward selection using AIC
MbckAIC <- step(object = Mfull, scope = list(lower = M0, upper = Mfull),
                trace = FALSE, direction = "backward", k = 2)
summary(MbckAIC)

## Prediction accuracy
MbckAIC.res <- pollutants$length[-sampleTrain] - predict(MbckAIC, newdata = dataTest)
mspeMbckAIC <- mean(MbckAIC.res^2)

print(paste("MSPE of foward selection model based on AIC:",mspeMbckAIC))
```

```{r}
## Backward selection using BIC 
MbckBIC <- step(object = Mfull, scope = list(lower = M0, upper = Mfull),
                trace = FALSE, direction = "backward", k = log(nrow(dataTrain)))
summary(MbckBIC)

## Prediction accuracy
MbckBIC.res <- pollutants$length[-sampleTrain] - predict(MbckBIC, newdata = dataTest)
mspeMbckBIC <- mean(MbckBIC.res^2)

print(paste("MSPE of foward selection model based on AIC:",mspeMbckBIC))
```

Correlation testing for PCB 1,2,4,5

```{r}
pollutants_pcbs <- dataTrain[,-c(2,3,5,6)]
cor(pollutants_pcbs[,2:8])
```

Manual analysis of variate correlation. This is done solely against the training set to avoid leakage of information from the holdout set, which would negatively impact the accuracy of model performance evaluation.
```{r, fig.align="center", fig.width=8, fig.height=9}
# Correlation matrix
dataTrain_num <- pollutants_original[row.names(dataTrain),]
dataTrain_cor <- cor(dataTrain_num[!colnames(dataTrain_num) %in% c('length')])

# Correlation plot
corrplot(dataTrain_cor, method="color", type="full", order="hclust",  tl.col = "black")
```

Based on the correlation plot, the following variates are correlated with other variates:PCB 1,3,4,5,8,9,10, furan 1, dioxin 1,2, Lymphocyte_pct. We remove them from the dataset and analyze the resulting correlation plot.
```{r, fig.align="center", fig.width=6, fig.height=6}
# New datasets
cols_to_exclude <- c('POP_PCB1','POP_PCB3','POP_PCB4','POP_PCB5','POP_PCB8',
                     'POP_PCB9','POP_PCB10','POP_furan1','POP_dioxin1',
                     'POP_dioxin2','lymphocyte_pct')
pollutants_corRemoved <- pollutants[!colnames(pollutants) %in% cols_to_exclude]
dataTrain_corRemoved <- dataTrain[!colnames(dataTrain) %in% cols_to_exclude]

# Correlation matrix
dataTrain_corRemoved_num <- pollutants_original[
  sampleTrain,
  !colnames(pollutants_original) %in% cols_to_exclude]

dataTrain_corRemoved_cor <- cor(
  dataTrain_corRemoved_num[
    !colnames(dataTrain_corRemoved_num) %in% c('length')])

# Correlation plot
corrplot(dataTrain_corRemoved_cor, method="color", type="full", order="hclust",
         tl.col = "black")
```

Stepwise model selection with PCB, Furan, Dioxin, and other correlated features removed

```{r}
# From Discord:
# Doing this [setting dataTrain2<-dataTrain, etc.] because we just did the manual variable selection based on the first training set
# For now, will just set dataTrain2 <- dataTrain and dataTest2 <- dataTest
# Later we can go through and remove all the 2's. I just don't want any conflicts rn in case anyone was modifying the second half of the analysis.

# Originally, this section defined a second training/test set. Changed this
set.seed(57)

N2 <- N
sampleTrain2 <- sampleTrain

dataTrain2 <- dataTrain
dataTest2 <- dataTest
```

Bounds for model selection

```{r}
M0_corRemoved <- lm(length ~ 1, data = dataTrain2)
Mfull_corRemoved <- lm(length ~ ., data = dataTrain2)
```

```{r}
## Forward selection using AIC
MfwdAIC_corRemoved <- step(object = M0_corRemoved, scope = list(lower = M0_corRemoved, upper = Mfull_corRemoved), 
             trace = FALSE, direction = "forward", k = 2)
summary(MfwdAIC_corRemoved)

## Prediction accuracy
MfwdAIC_corRemoved.res <- pollutants$length[-sampleTrain2] - predict(MfwdAIC_corRemoved, newdata = dataTest2)
mspeMfwdAIC_corRemoved <- mean(MfwdAIC_corRemoved.res^2)

print(paste("MSPE of foward selection model based on AIC:",mspeMfwdAIC_corRemoved))
```

```{r}
## Forward selection using BIC
MfwdBIC_corRemoved <- step(object = M0_corRemoved, scope = list(lower = M0_corRemoved, upper = Mfull_corRemoved), 
             trace = FALSE, direction = "forward", k = log(nrow(dataTrain2)))
summary(MfwdBIC_corRemoved)

## Prediction accuracy
MfwdBIC_corRemoved.res <- pollutants$length[-sampleTrain2] - predict(MfwdBIC_corRemoved, newdata = dataTest2)
mspeMfwdBIC_corRemoved <- mean(MfwdBIC_corRemoved.res^2)

print(paste("MSPE of foward selection model based on AIC:",mspeMfwdBIC_corRemoved))
```

```{r}
## Backward selection using AIC 
MbckAIC_corRemoved <- step(object = Mfull_corRemoved, scope = list(lower = M0_corRemoved, upper = Mfull_corRemoved),
                trace = FALSE, direction = "backward", k = 2)
summary(MbckAIC_corRemoved)

## Prediction accuracy
MbckAIC_corRemoved.res <- pollutants$length[-sampleTrain2] - predict(MbckAIC_corRemoved, newdata = dataTest2)
mspeMbckAIC_corRemoved <- mean(MbckAIC_corRemoved.res^2)

print(paste("MSPE of foward selection model based on AIC:",mspeMbckAIC_corRemoved))
```

```{r}
## Backward selection using BIC 
MbckBIC_corRemoved <- step(object = Mfull_corRemoved, scope = list(lower = M0_corRemoved, upper = Mfull_corRemoved),
                trace = FALSE, direction = "backward", k = log(nrow(dataTrain2)))
summary(MbckBIC_corRemoved)

## Prediction accuracy
MbckBIC_corRemoved.res <- pollutants$length[-sampleTrain2] - predict(MbckBIC_corRemoved, newdata = dataTest2)
mspeMbckBIC_corRemoved <- mean(MbckBIC_corRemoved.res^2)

print(paste("MSPE of foward selection model based on AIC:",mspeMbckBIC_corRemoved))
```
























